---
author: "Anusri"
date: "2025-02-10"
output: html_document
---

```{r}
haversine <- function(lat1, lon1, lat2, lon2) {
  r <- 3959  # Radius of Earth in miles
  lat1 <- lat1 * pi / 180
  lon1 <- lon1 * pi / 180
  lat2 <- lat2 * pi / 180
  lon2 <- lon2 * pi / 180
  dlat <- lat2 - lat1
  dlon <- lon2 - lon1
  a <- sin(dlat / 2)^2 + cos(lat1) * cos(lat2) * sin(dlon / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  return(r * c)
}

```

I have defined the Haversine function which calculates the great-circle distance between two latitude-longitude points 

```{r}
# Load necessary library
install.packages("readxl")
library(readxl)
```

Installing the required packages

```{r}
# Load the dataset
file_path <- ("C:/Users/anusr/Downloads/clinics.xls")  
df <- read_excel(file_path)

# Convert to a data frame
df <- as.data.frame(df)
```
Loaded the dataset

```{r}
# For-loop implementation
haversine_looping <- function(df) {
  distances <- numeric(nrow(df))
  for (i in 1:nrow(df)) {
    distances[i] <- haversine(40.671, -73.985, df$locLat[i], df$locLong[i])
  }
  return(distances)
}


str(df)  # Checking the structure of the dataframe
df$locLat <- as.numeric(df$locLat)
df$locLong <- as.numeric(df$locLong)
df <- df[!is.na(df$locLat) & !is.na(df$locLong), ]  # Removing rows with NA values
df$distance <- haversine(40.671, -73.985, df$locLat, df$locLong)


# Measure execution time
library(microbenchmark)
microbenchmark(for_loop = haversine_looping(df), times = 1)

```
I have implemented for loop which calculates distances iterating through each row of the dataframe. Converted latitude and longitude columns to numeric and removes NA values before computation. The execution time of the for-loop approach is measured using the microbenchmark library.And I got min of 5.5907milliseconds. This method is less efficient than vectorized approaches but helps in understanding the process.



```{r}
#Using apply() method
haversine_apply <- function(df) {
  distances <- apply(df, 1, function(row) {
    haversine(40.671, -73.985, row["locLat"], row["locLong"])
  })
  return(distances)
}


str(df)  
summary(df)  
df$locLat <- as.numeric(df$locLat)
df$locLong <- as.numeric(df$locLong)
df <- df[!is.na(df$locLat) & !is.na(df$locLong), ]
haversine_apply <- function(df) {
  distances <- apply(df, 1, function(row) {
    lat <- as.numeric(row["locLat"])
    lon <- as.numeric(row["locLong"])
    if (is.na(lat) | is.na(lon)) return(NA)  # Skip invalid rows
    haversine(40.671, -73.985, lat, lon)
  })
  return(distances)
}

df$distance <- haversine_apply(df)


# Measure execution time
microbenchmark(apply_method = haversine_apply(df), times = 1)

```
For one more way I used apply() function.It ensures latitude and longitude values are numeric and removes NA values before computation. Compared to a for-loop, this approach is more efficient but still slower than vectorized operations and the execution time of the apply-based approach is measured using the microbenchmark library. I got min of 5.439.

```{r}
#Using Vectorized approach
haversine_vectorized <- function(df) {
  r <- 3959  # Radius of Earth in miles
  lat1 <- 40.671 * pi / 180
  lon1 <- -73.985 * pi / 180
  lat2 <- df$locLat * pi / 180
  lon2 <- df$locLong * pi / 180
  dlat <- lat2 - lat1
  dlon <- lon2 - lon1
  a <- sin(dlat / 2)^2 + cos(lat1) * cos(lat2) * sin(dlon / 2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  return(r * c)
}

# Measure execution time
microbenchmark(vectorized = haversine_vectorized(df), times = 1)

```
Vectorized function computes distances using a fully vectorized approach, leveraging element-wise operations.This avoids explicit loops, making it significantly faster than both the for-loop and apply methods. The vectorized implementation directly computes distances for all rows in a single operation and execution time is measured using the microbenchmark library to compare efficiency.


```{r}
#Tabulating the results
results <- microbenchmark(
  for_loop = haversine_looping(df),
  apply_method = haversine_apply(df),
  vectorized = haversine_vectorized(df),
  times = 1
)

print(results)

```



